---
title: "Descriptive statistics"
author:
  - name: "Sam Nooij"
    affiliations:
      - ref: UU-KLIF
    orcid: 0000-0001-5892-5637
    email: s.nooij [at] uu.nl
    corresponding: true

affiliations:
  - id: UU-KLIF
    name: Utrecht University
    department: Faculty of Veterinary Medicine, Department of Clinical Infectiology
    address: Yalelaan 1
    city: Utrecht
    country: The Netherlands
    postal-code: 3584CL

date: "2024-12-03"

execute: 
  echo: true
  warning: false
  message: false
  cache: true

format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    fig-height: 4
    fig-width: 7
  pdf:
    documentclass: report
    classoption: [twocolumn, landscape]
    mainfont: "Open Sans"
    mainfontoptions:
      - Color=101010
    colorlinks: true
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 3
    highlight-style: github

knitr:
  opts_chunk: 
    cache: true

editor:
  markdown:
    wrap: 72
---

# CRISPR spacers in *Campylobacter*

**Descriptive statistics of CRISPR arrays in *Campylobacter* genomes
from
[AllTheBacteria](https://allthebacteria.readthedocs.io/ "AllTheBacteria documentation")**

We have downloaded all the *Campylobacter* *jejuni* and *C. coli*
genomes from AllTheBacteria (ATB) to screen for CRISPR spacers
(@fig-number_of_species). The current experiment includes all genomes
that have been classified as 'Campylobacter_D jejuni\*' or
'Campylobacter_D coli\*' using
[Sylph](https://www.nature.com/articles/s41587-024-02412-y "Shaw & Yu, 2024. Nature Biotechnology")
(which the creators of ATB ran on all genomes), that have been added to
ATB in the incremental update of August 2024. (∗: any species
designation matching these name prefixes were included.) These genomes
have been screened for CRISPR ([Clustered Regularly Interspaced Short
Palindromic
Repeats](https://en.wikipedia.org/wiki/CRISPR "CRISPR wiki")) spacers
and *Cas* (CRISPR-associated) genes using
[CCTyper](https://www.biorxiv.org/content/10.1101/2020.05.15.097824v1 "Russel et al., 2020. bioRxiv")
(version 1.8.0). CCTyper produces tab-separated tables that summarise
the CRISPR spacers, the repeat sequences in between, Cas genes including
statistics and sequences of the spacers.

```{r setup, include=FALSE}
library(knitr)
library(rmarkdown)
library(here)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(viridisLite)
library(patchwork) # Combine figures while aligning axes
library(rstatix) # for convenient statistical tests and p-value reporting
library(scales) # for manually adjusting log-scale figure axes
library(colorspace) # for filling boxplots/violins with lighter shade of same colour

knitr::opts_knit$set(base.dir = here(""))
```

In this analysis report, we summarise the prevalence and other general
descriptive statistics of the CRISPR spacers. Therefore, we first need
to import the results from CCTyper:

```{r import_cctyper_results}
#| message: false

# Files were originally processed in batches
batches <- Sys.glob("../data/tmp/cctyper/batch_*")

# Read the tab-separated files for each batch and concatenate them
read_multiple_tables <- function(input_files) {
  return(
    data.frame(Filename = input_files,
               read_delim(input_files,
                          show_col_types = F,
                          trim_ws = T))
  )
}

# Try reading concatenated files, or read separate files from batches
#  and save as one concatenated file
read_or_create <- function(table_file) {
  if(file.exists(table_file)) {
    message(paste(basename(table_file), "exists already, reading from file!"))
    return(read_delim(table_file, show_col_types = F))

  } else {
    message(paste(
      basename(table_file),
      "does not yet exist. Concatenating batch tables..."))

    table_wildcard <- gsub(x = basename(table_file),
                           pattern = "concatenated",
                           replace = "*")
    file_list <- Sys.glob(paste0(batches, "/", table_wildcard))
    df <- do.call("rbind", lapply(file_list, read_multiple_tables))
    write_delim(x = df, file = table_file, delim = "\t")
    return(df)
  }
}


CRISPR_Cas_df <- read_or_create("../data/processed/CRISPR_Cas-concatenated.tab")

crisprs_all_df <- read_or_create("../data/processed/crisprs_all-concatenated.tab")

crisprs_near_cas_df <- read_or_create("../data/processed/crisprs_near_cas-concatenated.tab") %>%
  select(-Contig) %>%
  separate_wider_delim(cols = CRISPR, delim = ".",
                       names = c("sample_accession", "Contig")) %>%
  separate_wider_delim(cols = Contig, delim = "_",
                       names = c("Contig", "CRISPR_number"))

crisprs_orphan_df <- read_or_create("../data/processed/crisprs_orphan-concatenated.tab") %>%
  select(-Contig) %>%
  separate_wider_delim(cols = CRISPR, delim = ".",
                       names = c("sample_accession", "Contig")) %>%
  separate_wider_delim(cols = Contig, delim = "_",
                       names = c("Contig", "CRISPR_number"))
```

The output of CCTyper only contains sample accession IDs that can be
used to identify samples. Import a minimal set of ENA metadata and
species identifications by Sylph:

```{r import_metadata}
#| fig-width: 5
#| fig-height: 1
#| label: fig-number_of_species
#| fig-cap: "Number of species present in dataset"
#| message: false

campylobacter_metadata_file <- "../data/processed/Campylobacter_ATB_metadata.tsv"

if(file.exists(campylobacter_metadata_file)) {
  message("Reading Campylobacter metadata file!")
  campylobacter_metadata <- read_delim(
    campylobacter_metadata_file,
    show_col_types = F)

} else {
  message("Creating Campylobacter metadata file...")
  total_campylobacter_metadata <- read_delim(
    "../data/ATB/ena_metadata.20240801.selection-only_Campylobacter.tsv.gz",
    delim = "\t",
    col_names = F,
    show_col_types = F)

  older_campylobacter_metadata <- read_delim(
    "../data/ATB/ena_metadata.0.2.20240606.selection-only_Campylobacter.tsv.gz",
    delim = "\t",
    col_names = F,
    show_col_types = F)

  # Remove the older entries to keep only the Campylobacter genomes
  # from the incremental update
  campylobacter_metadata <- total_campylobacter_metadata %>%
    filter(! X2 %in% older_campylobacter_metadata$X1)

  # Take the column names from the table before grepping
  colnames(campylobacter_metadata) <- read_delim(
    file = "../data/ATB/ena_metadata.20240801.selection.tsv.gz",
    n_max = 1,
    delim = "\t",
    show_col_types = F
  ) %>% colnames()

  # Remove columns that contain only NA values
  campylobacter_metadata <- campylobacter_metadata %>%
    select_if(~sum(!is.na(.)) > 0)

  simplified_species <- read_delim(
    file = "../data/ATB/sylph.tsv.gz",
    delim = "\t",
    show_col_types = F) %>%
    select(Sample, Species)

  campylobacter_metadata <- left_join(
    x = campylobacter_metadata,
    y = simplified_species,
    by = c("sample_accession" = "Sample")
  )

  # Remove all genomes whose taxonomic annotation (species) does
  # not start with 'Campylobacter'
  campylobacter_metadata <- campylobacter_metadata %>%
    filter(grepl("Campylobacter_D jejuni|Campylobacter_D coli", Species))

  write_delim(x = campylobacter_metadata,
              file = campylobacter_metadata_file,
              delim = "\t")
}

campylobacter_species_numbers <- campylobacter_metadata %>%
  mutate(species = if_else(grepl("jejuni", Species), "jejuni", "coli")) %>%
  count(species)

# Alternative visualisation method: stacked bar chart
stacked_species_plot <- ggplot(data = campylobacter_species_numbers,
                             aes(x = 0, y = n, fill = species)) +
  geom_bar(stat = "identity") +
  theme_pubr() +
  ggtitle(paste("Total genomes:",
                format(sum(campylobacter_species_numbers$n),
                       big.mark = ","))) +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "right") +
  scale_fill_manual(name = "Species",
                    values = c("#895d45", "#c6903c")) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  geom_text(aes(label = n %>%
                  format(big.mark = ",")),
            position = position_stack(vjust = 0.5),
            size = 4.5,
            color = "black")

stacked_species_plot
```

# Spacers per genome

We have identified CRISPR spacer in all *Campylobacter jejuni* and *C.
coli* genomes using
[CCTyper](https://github.com/Russel88/CRISPRCasTyper "GitHub repository for CRISPRCasTyper").
CCTyper separately annotates CRISPR spacers near *Cas* genes and
separate CRISPR arrays, called 'orphans'. To inventory the prevalence
and size of both types of CRISPR arrays, show for each genome the number
of CRISPR spacers as scatterplot (@fig-crisprs_per_genome).

```{r spacers_per_genome}
#| label: fig-crisprs_per_genome
#| fig-cap: "Number of CRISPR spacers per genome"
#| fig-height: 4.5
#| fig-width: 7.5

annotated_crispr_number_df <- left_join(
  x = campylobacter_metadata %>%
    select(sample_accession, country, host, isolation_source,
           scientific_name, tax_id, Species, description),
  y = full_join(
    x = crisprs_near_cas_df %>%
      group_by(sample_accession) %>%
      count() %>%
      rename("CRISPRs_near_Cas" = "n"),
    y = crisprs_orphan_df %>%
      group_by(sample_accession) %>%
      count() %>%
      rename("Orphan_CRISPRs" = "n"),
    by = "sample_accession"
  )
) %>%
  mutate(
    across(CRISPRs_near_Cas:Orphan_CRISPRs,
           ~replace_na(., 0))
  )

number_of_crisprs_scatterplot <- ggplot(
  data = annotated_crispr_number_df,
  mapping = aes(x = CRISPRs_near_Cas,
                y = Orphan_CRISPRs)
) +
  geom_point(position = position_jitter(),
             size = 0.7, alpha = 0.3) +
  theme_pubr()

crispr_near_cas_density = ggplot(
  data = annotated_crispr_number_df,
  mapping = aes(x = CRISPRs_near_Cas)) +
  geom_density(fill = "grey20") +
  theme_pubr() +
  theme(axis.title.x = element_blank())

orphan_crispr_density = ggplot(
  data = annotated_crispr_number_df,
  mapping = aes(y = Orphan_CRISPRs)) +
  geom_density(fill = "grey20") +
  theme_pubr()

design <- "
1111#
22223
22223
"

spacers_per_genome_figure <- crispr_near_cas_density +
  number_of_crisprs_scatterplot +
  orphan_crispr_density +
  plot_layout(design = design,
              axes = "collect")

spacers_per_genome_figure
```

From this figure, it appears that by far the most genomes have no CRISPR
spacers at all, and otherwise they mostly have one CRISPR spacer near
*Cas* genes. At most, CCTyper has found 3 CRISPRs near *Cas* genes and
up to 4 orphan CRISPR spacers. Let's see this distribution in numbers
and separate them between *C. jejuni* and *coli*.

```{r spacers_per_genome_stats}
total_genomes = nrow(annotated_crispr_number_df)

annotated_crispr_number_df %>%
  count(CRISPRs_near_Cas, name = "Genomes") %>%
  mutate(Percentage = round(Genomes / total_genomes * 100, 2))

print("CRISPRs in Campylobacter jejuni")
annotated_crispr_number_df %>%
  filter(grepl(pattern = "Campylobacter_D jejuni", Species)) %>%
  group_by(CRISPRs_near_Cas) %>%
  summarise(Genomes = length(CRISPRs_near_Cas)) %>%
  mutate(total_genomes = sum(Genomes),
         Percentage = round(Genomes / total_genomes * 100, 2)) %>%
  select(-total_genomes)
# The number of other jejuni subspecies(?) is too low to do
# separate analyses, so just put them all together

print("Orphan CRISPRs in C. jejuni")
annotated_crispr_number_df %>%
  filter(grepl(pattern = "Campylobacter_D jejuni", Species)) %>%
  group_by(Orphan_CRISPRs) %>%
  summarise(Genomes = length(Orphan_CRISPRs)) %>%
  mutate(total_genomes = sum(Genomes),
         Percentage = round(Genomes / total_genomes * 100, 2)) %>%
  select(-total_genomes)

print("CRISPRs in Campylobacter coli")
annotated_crispr_number_df %>%
  filter(grepl(pattern = "Campylobacter_D coli", Species)) %>%
  group_by(CRISPRs_near_Cas) %>%
  summarise(Genomes = length(CRISPRs_near_Cas)) %>%
  mutate(total_genomes = sum(Genomes),
         Percentage = round(Genomes / total_genomes * 100, 2)) %>%
  select(-total_genomes)

print("Orphan CRISPRs in C. coli")
annotated_crispr_number_df %>%
  filter(grepl(pattern = "Campylobacter_D coli", Species)) %>%
  group_by(Orphan_CRISPRs) %>%
  summarise(Genomes = length(Orphan_CRISPRs)) %>%
  mutate(total_genomes = sum(Genomes),
         Percentage = round(Genomes / total_genomes * 100, 2)) %>%
  select(-total_genomes)
```
